\LaTeX\ Technical Documents with Docker and Make

### 2020-12-08 ms.tex
various major changes

\author{Paschalis~Bizopoulos and Dimitris~Bizopoulos
\thanks{P. Bizopoulos and D. Bizopoulos are Independent Researchers, Thessaloniki, Greece e-mail: pbizopoulos@protonmail.com, dimitrisbizopoulos@gmail.com}}

@@

\begin{abstract}
	We propose a template for writing portable \LaTeX\ technical documents with Docker and \textit{Make} in POSIX-oriented operating systems.
	The template provides a Makefile that allows the author to execute the code that generates the figures, tables and variables (results), which are then used during the \LaTeX\ compilation, to produce the fast/draft and slow/final versions of the document.
	We release an open source repository of the template and use it to write this paper.
\end{abstract}

\section{Introduction}
There are multiple ways for writing technical documents that present both computationally generated results (figures, tables and variables) and natural text.
One programming paradigm that tries to solve this task is \textit{literate programming}~\cite{knuth1984literate} in which the author inserts snippets of code and its executed output alongside natural text.
Some applications of \textit{literate programming} include Jupyter~\cite{kluyver2016jupyter} which allows publishing code, results and explanations in a executable format, PythonTeX~\cite{poore2015pythontex} which allows Python to be executed in \LaTeX\ and ActivePapers~\cite{hinsen2014activepapers} in which code can be executed using the Java Virtual Machine.

@@

We propose a template for writing portable \LaTeX\ technical documents with Docker and \textit{Make} in POSIX-oriented operating systems and provide an open source implementation\footnote{\url{https://github.com/pbizopoulos/latex-technical-documents-with-docker-and-make-template}}.
We write this paper\footnote{\url{https://arxiv.org/abs/2005.12660}}\footnote{\url{https://github.com/pbizopoulos/latex-technical-documents-with-docker-and-make}} using this template and choose the Python programming language~\cite{van2007python} for generating the results.

For the rest of the paper we will refer to:
\begin{itemize}
	\item \textit{results} as the figures, tables and variables that are generated from a computation,
	\item \textit{results code} as the scientific-oriented programming language code (\textit{main.py}) that produces the \textit{results},
	\item \textit{\LaTeX\ code} as the natural text (\textit{ms.tex}) and bibliography (\textit{ms.bib}),
	\item \textit{code} as both the \textit{\LaTeX\ code} and \textit{results code},
	\item \textit{document} as the technical document that contains the rendered natural text and \textit{results} and
	\item \textit{author} as the author of the \textit{document} that writes and develops the \textit{code}.
\end{itemize}

@@

The template makes use of the automatic generation of intermediate files of the \textit{Make} program.
The \textit{author} only needs to edit (or use the \textit{touch} Unix utility to update the modification time) a file from \textit{results code} or \textit{\LaTeX\ code} and the corresponding compilation is automatically triggered with \textit{Make}.

@@

Setting the random seeds and the previously mentioned \LaTeX\ packages we could have deterministic stochasticity that generates figures, tables and variables.

@@

In cases where reproducibility is required, the potential differences in the resulting \textit{document} binary can be identified using tools such as diffoscope\footnote{\url{https://diffoscope.org/}} and the cache files saved at the tmp directory after execution.

@@

\section{Discussion}
The proposed template is based on plain text files and can be used with any POSIX-oriented operating that supports Docker and \textit{Make}.
Moreover it can be combined with any version control system, docker container registry providor and text editor thus preventing `app/vendor lock-in' situations.

\section{Conclusion}
Creation, usage and publication of technical documents are one of the top challenges for reproducible technical documents~\cite{barba2019praxis}.
The proposed template aids to this purpose in writing reproducible technical documents using the tools that most of academics already use (\LaTeX, Docker).

>>>>

\author{Paschalis~Bizopoulos
\thanks{P. Bizopoulos is an Independent Researcher, Thessaloniki, Greece e-mail: pbizopoulos@protonmail.com}}

@@

\begin{abstract}
	We propose a template for writing portable \LaTeX\ technical documents with Docker and Make in POSIX-oriented operating systems.
	The template provides a Makefile that allows the author to execute the code that generates the variables, tables and figures (results), which are then used during the \LaTeX\ compilation, to produce the fast/draft and slow/final versions of the document.
	We release an open source repository of the template and use it to write this paper.
\end{abstract}

\section{Introduction}
There are multiple ways for writing technical documents that present both computationally generated results (variables, tables and figures) and natural text.
One programming paradigm that tries to solve this task is \textit{literate programming}~\cite{knuth1984literate} in which the author inserts snippets of code and its execution output alongside natural text.
Some applications of \textit{literate programming} include Jupyter~\cite{kluyver2016jupyter} which allows publishing code, results and explanations in a executable format, PythonTeX~\cite{poore2015pythontex} which allows Python to be executed in \LaTeX\ and ActivePapers~\cite{hinsen2014activepapers} in which code can be executed using the Java Virtual Machine.

@@

We propose a template for writing portable \LaTeX\ technical documents with Docker and Make in POSIX-oriented operating systems and provide an open source implementation\footnote{\url{https://github.com/pbizopoulos/latex-technical-documents-with-docker-and-make-template}}.
We write this paper\footnote{\url{https://github.com/pbizopoulos/latex-technical-documents-with-docker-and-make}}\footnote{\url{https://arxiv.org/abs/2005.12660}} using this template and choose the Python programming language~\cite{van2007python} for generating the results.

For the rest of the paper we will refer to:
\begin{itemize}
	\item \textit{results code} as the scientific-oriented programming language code (\textit{main.py}) that produces the variables, tables and figures,
	\item \textit{results} as the variables, tables and figures that are generated by the \textit{results code},
	\item \textit{\LaTeX\ code} as the natural text (\textit{ms.tex}) and bibliography (\textit{ms.bib}),
	\item \textit{code} as both the \textit{\LaTeX\ code} and \textit{results code},
	\item \textit{document} as the technical document that contains the rendered natural text and \textit{results} and
	\item \textit{author} as the author of the \textit{document} that writes and develops the \textit{code}.
\end{itemize}

@@

The template makes use of the automatic generation of intermediate files of the Make program.
The \textit{author} only needs to edit (or use the \textit{touch} POSIX utility to update the modification time) a file from \textit{results code} or \textit{\LaTeX\ code} and the corresponding compilation is automatically triggered with Make.

@@

Setting the random seeds and the previously mentioned \LaTeX\ packages we could have deterministic stochasticity that generates variables, tables and figures.

@@

In cases where reproducibility checks are required, the POSIX utility \textit{cmp} can be used.
For specifically identifying the sources of non-determinism an author can use tools such as diffoscope\footnote{\url{https://diffoscope.org/}} in the resulting \textit{document} binaries as follows:
\begin{lstlisting}[language=bash, style=lststyle, caption={Test fast/draft document reproducibility.}]
# Generate the first fast/draft document version
make
# Backup the first resulting pdf
cp tmp/ms.pdf tmp/ms-previous.pdf
# Update the modification timestamp of main.py
touch main.py
# Generate the second fast/draft document version
make
# Compare fast/draft document versions byte by byte
cmp tmp/ms.pdf tmp/ms-previous.pdf
# Compare fast/draft document versions
# in a human readable form
diffoscope tmp/ms.pdf tmp/ms-previous.pdf
\end{lstlisting}

@@

\section{Discussion}
The proposed template is based on plain text files and can be used with any POSIX-oriented operating that supports Docker and Make.
Moreover it can be combined with any version control system, docker container registry providor and text editor thus preventing `app/vendor lock-in' situations.

\section{Conclusion}
Creation, usage and publication of technical documents are one of the top challenges for reproducible technical documents~\cite{barba2019praxis}.
The proposed template aids to this purpose in writing reproducible technical documents using the tools that most of academics already use (\LaTeX, Docker, Make).
